---
title: "Untitled"
author: "jbot"
date: "09/07/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("DESeq2")
library(tidyverse)
library(ggpubr)
library(magrittr)
```

# loading data

loading counts table

```{r}
#loading data
counts <- data.frame(read_tsv("NGEI044_exon_featureCounts.raw"))

#renaming gene id
counts <- counts %>% 
  dplyr::rename(gene_id = Geneid)

#saving columns that will be removed
row_info <- counts %>%
  dplyr::select(gene_id:gene_name)

#removing columns that belong in row info
counts <- counts %>% 
  dplyr::select(-(Chr:gene_name))

#adding rownames
rownames(counts) <- counts$gene_id

#removing gene id column
counts <- counts %>%
  dplyr::select(-gene_id)
```

loading exp design matrix

```{r}
exp_design <- read.csv("exp_design.csv", row.names=1, sep = ";")

#making factors
exp_design$protein  <- as.factor(exp_design$protein)
exp_design$guide  <- as.factor(exp_design$guide)
exp_design$targeting  <- as.factor(exp_design$targeting)

#setting untrt as first level
exp_design$protein %<>% relevel("untrt")
exp_design$guide %<>% relevel("untrt")
exp_design$targeting %<>% relevel("untrt")

exp_design
```

Construct the DESeqDataSet object

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts,
                                 colData = exp_design,
                                 design = ~ protein + guide + targeting)
```

# Exploratory analysis and visualization

## Pre-filtering

```{r}
keep <- rowSums(counts(dds)) >= 10
dds_fil <- dds[keep,]
```

## Transform the counts for PCA.

```{r}
vsd <- vst(dds_fil, blind = FALSE)
head(assay(vsd), 3)
colData(vsd)
```

## PCA plot

```{r}
library(ggrepel)
pca_data <- plotPCA(vsd, intgroup = c("protein", "guide", "targeting"), returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2)) +
  geom_point(size = 3) +
  geom_text_repel(label = pca_data$name) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA with VST data")

#pca_data
```

#Differential expression analysis

## Running the differential expression pipeline

```{r}
dds <- DESeq(dds_fil)
```

## Building the results tables

FDR of 5%

```{r}
res_protein <- results(dds, contrast=c("protein","trt","untrt"), alpha = 0.05, lfcThreshold = 1)
res_guide <- results(dds, contrast=c("guide","trt","untrt"), alpha = 0.05, lfcThreshold = 1)
res_targeting <- results(dds, contrast=c("targeting","trt","untrt"), alpha = 0.05, lfcThreshold = 1)
```

Protein summary
```{r}
summary(res_protein)
```

Guide summary
```{r}
summary(res_guide)
```

Targeting summary
```{r}
summary(res_targeting)
```

## MA-plots

```{r}
library("apeglm")
```

```{r}
MA_protein <- lfcShrink(dds, coef="protein_trt_vs_untrt", type="apeglm")
MA_guide <- lfcShrink(dds, coef="guide_trt_vs_untrt", type="apeglm")
MA_targeting <- lfcShrink(dds, coef="targeting_trt_vs_untrt", type="apeglm")
```

For protein

```{r}
DESeq2::plotMA(MA_protein, ylim = c(-5,5), alpha = 0.05)
```

For guide

```{r}
DESeq2::plotMA(MA_guide, ylim = c(-5, 5), alpha = 0.05)
```

For targeting

```{r}
DESeq2::plotMA(MA_targeting, ylim = c(-7, 7), alpha = 0.05)
```

#targeted

```{r}
library(AnnotationDbi)
library("org.Hs.eg.db")
```

```{r}
ens.str.tar <- substr(rownames(res_targeting), 1, 15)
res_targeting$symbol <- mapIds(org.Hs.eg.db,
                     keys=ens.str.tar,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
res_targeting$targetingentrez <- mapIds(org.Hs.eg.db,
                     keys=ens.str.tar,
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
resOrdered <- res_targeting[order(res_targeting$pvalue),]
head(resOrdered)
```

```{r}
gene_id <- rownames(res_targeting)
up_targeting <- res_targeting %>%
  as.data.frame() %>%
  cbind(gene_id) %>%
  filter(padj < 0.05) %>%
  filter(log2FoldChange > 1) %>%
  mutate(tenlogp = -log10(padj))

#write.csv2(up_targeting, file = "up_targeting.csv")
```

```{r}
gene_id <- rownames(res_targeting)
down_targeting <- res_targeting %>%
  as.data.frame() %>%
  cbind(gene_id) %>%
  filter(padj < 0.05) %>%
  filter(log2FoldChange < -1) %>%
  mutate(tenlogp = -log10(padj))

#write.csv(down_targeting, file = "down_targeting.csv")
```

# Volcano plot

```{r}
library('EnhancedVolcano')
gene_id <- rownames(res_targeting)
```


```{r}
volcano_plot <- EnhancedVolcano(res_targeting,
    lab = res_targeting$symbol,
    selectLab = c(#upregulated
                  "NRARP", "ARC", "CDKN1A", "LIF", "EGR2", "FOS", "KLF6", "IL11", "CDKN2B", "INHBA",
                  "CXCL8", "JUNB", "DUSP5", "DUSP1", "DUSP10", "DUSP8", "IGF2", "EGR1", "IL16", "CXCL11",
                  "CXCL3", "CXCL2", "JUN", "IL24", "ATF3", "KLF4", "CXCL10", "DUSP28", "EGR3", "EGR4", 
                  "C6orf141", "LINC00452",
                  #downregulated
                  "ATP6", "ATP8", "FBLN1", "BCAT1"),
    x = 'log2FoldChange',
    y = 'pvalue',
    pCutoff = 0.05,
    FCcutoff = 1,
    title = element_blank(),
    subtitle = element_blank(),
    legendPosition = "right",
    drawConnectors = TRUE,
    widthConnectors = 0.25,
    labSize = 1.5) #+
    #coord_flip()
volcano_plot
```

```{r}
volcano_plot +
  theme_pubr(base_size = 8) +
  theme(legend.position = "right",
        text = element_text(family = "sans")) -> volcano_plot
volcano_plot
```


```{r}
ggsave(filename = "volcano_plot_select.pdf", 
       volcano_plot, 
       width = 110, 
       height = 65,
       useDingbats=FALSE,
       units = "mm")
```
